---
title: README
author: Michael Weaver
date: December 20, 2024
engine: knitr
output:
  md_document:
    variant: gfm
editor_options: 
  chunk_output_type: inline
---

```{r include=FALSE}
devtools::load_all()
```

# swampBioaccumulationAnalysis

R package to analyze SWAMP Bioaccumulation Monitoring Data

## Installation

You can install the **swampBioaccumulationAnalysis** package from GitHub.

```{r eval=FALSE}
install.packages("remotes") # if you have not installed "remotes" package
remotes::install_github("CAWaterBoardDataCenter/swamp-bioaccumulation-analysis")
```

The source code for **swampBioaccumulationAnalysis** package is available on GitHub at

-   <https://github.com/CAWaterBoardDataCenter/swamp-bioaccumulation-analysis>.

## Usage

The **swampBioaccumulationAnalysis** package provides several functions that can be used to analyze your dataset.

-   `summary_tables()`
-   `length_adjusted_averages()`
-   `mercury_by_species_barplot()`
-   `mercury_vs_length_scatterplot()`
-   `all_analyses()`

To use these functions, first, you should load the package as follows:

```{r eval=FALSE}
library(swampBioaccumulationAnalysis)
```

### Data Source Requirements

**swampBioaccumulationAnalysis** is designed to analyze tissue datasets downloaded from the [CEDEN Advanced Query Tool](https://ceden.waterboards.ca.gov/Home/virtAqtTool), in .tsv file format, and any source files evaluated by package functions will need to be in that file format and data stucture.

It’s recommended that users place the .tsv file downloaded from CEDEN AQT download in the folder that they have set as their R working directory, so they can run the analysis functions simply using the filename for the functions' filename argument. (Otherwise they will need to pass the dataset file’s filepath relative to their R working directory, or its absolute filepath on their machine, as the filename argument.)

For example, if your dataset was in a file named "CEDENDownload.tsv" in the folder `C:\Users\yourname\Documents\` on your local machine, you could set your R working directory by running the following in your R console:

```{r eval=FALSE}
setwd("C:\Users\yourname\Documents")
```

and then you could run a function from the package on your dataset as follows:

```{r eval=FALSE}
summary_tables("CEDENDownlaod.tsv")
```

### Function Details

  * `summary_tables(`*`filename`*`)` Outputs a six tab (.xlsx) spreadsheet with summary tables for your dataset:
    * Analytes included in the dataset, detection limits, number of observations, and frequencies of detection and reporting (in tab *AnalytesObservsDetects*)
    * Mercury in individual sport fish (in tab *SportfishMercuryIndiv*)
    * Summary of sport fish results, with composites or means at each location (in tab *SportfishCompsAndMeans*)
    * Summary of prey fish results (in tab *PreyfishComposites*)
    * Scientific and common names of prey fish species in the dataset, the number of locations in which they were sampled, and their minimum, median, and maximum total lengths (mm) (in tab *PreyfishCounts*)
    * Scientific and common names of sport fish species in the dataset, the number of locations in which they were sampled, numbers of individual or composite samples, their minimum, median, and maximum total lengths (mm), and whether they were analyzed as composites or individuals (in tab *SportfishCounts*)
* `length_adjusted_averages(`*`filename`*`)` Outputs a three tab (.xlsx) spreadsheet with length-adjusted averages (in tab *LengthAdjustedAverages*), length-adjusted averages with additional summary calculations (in tab *LengthAdjustedAveragesDetail*), and intermediate length-adjusted result calculations (in tab *SupportingCalculations*)
* `mercury_by_species_barplot(`*`filename`*`,`*`fish_class`*`)` Outputs a (.jpeg file) bar plot for mercury concentration by species, for a given source file and type of fish (sport or prey)
* `mercury_vs_length_scatterplot(`*`filename`*`,`*`species_common_name`*`)` Outputs a (.jpeg file) scatterplot of fish mercury concentration vs fish length in (mm) for a given source file and species
* `all_analyses(`*`filename`*`)` Runs all of the other functions in the package on your source file. (The *mercury_by_species_barplot()* function is run for both sport and prey fish. The *mercury_vs_length_scatterplot()* function is run for every species present in the dataset with mercury results.) **If you would like to generate the full set of tables and figures that the package can create for your dataset, you can do so solely by running this one function.**

Once the package is installed and loaded, you can run `?functionname` or `help("functionname")` in the R Console to view full function-specific documentation

## Package Data

The package contains external data (.csv) files in the installation directory's extdata/ subdirectory that contain reference and lookup list information supporting the analysis functions. There are three data files:

- `mercury-units.csv`
    Contains unit conversions to ppm for reported units for concentration of mercury in fish tissue, in order for for the dataset's mercury in fish results to be compared with statewide water quality objectives. 
- `species-type-lookup-table.csv`
    Contains classifications of fish species (as prey, sport or both). Fish species classified as both include threshold values (for total fish length in mm) to determine whether individual fish in the dataset are treated as prey or sport fish.
- `summing-lookup.csv`
    Contains a crosswalk of individual analytes to summing groups, in order for the sum of results for an analyte group to be reported in place of the individual analyte results in some summary tables (e.g. "Sum of PCBs" in place of individual PCB congeners).

